Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-15 agujournaltemplate.tex
Update Supplement
Start rewrite for Ago/Andrew

\journalname{Enter journal name here}

@@

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
% \index Previous studies proposed a shift in the community, but with the full dataset we see a shift and a return. The community shows a departure form cluster one around 2004, but a return to that community in 2014. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind. 
\end{keypoints}

@@

@@

\begin{table}
\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\label{table:clustering}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & *\\
\addlinespace
Temperature & 4763.0 & 0.003 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
Salinity & 7636.0 & 0.002 & **\\
\addlinespace
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
\caption{Gradient Forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
\label{fig:GF}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
\caption{Correlation matrix with clustering of environmental variables, diversity metrics and cell counts. Correlation was calcualted using a Spearman's rank correlation coefficient, the magnitude of the coefficient is indicated by the color of the circle. The size of the circle and significance level marking represents the calculated significance of the correlation deviating from 0 (p-value: ***\textless0.001, **\textless0.01, *\textless0.05). The time series were aditionally ordered by hierarchical clustering via the complete linkage method into 5 groups indicated by the black square border.}
\label{fig:map}
\end{figure}

@@

% HERE ADD FIGURE OF GF RUN WITHOUT TIME LAGS (ALLVARS)

@@

\caption{Resulting $R^2$ weighted importance for running predictor variables together with time-lags to check if they provide a better fit to the data. Each column represents a single run, with resulting importance values for time series and monthly lags. The results were stable, so that repeated runs confirmed the chosen time lag. For in-situ variables,  we tested all measurements up to a lag of 3 months. For climate variables we tested up to a lag of 6 months and added a lag of 12 months for the climate indices. Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
lag & NO3 & PO4 & SiO4 & SST & \qty{21}{\celsius} Isotherm & Salinity & AMO & MEI v.2 & Wind speed & Precipitation & Evaporation\\
\midrule
0 & \textbf{0.027} & \textbf{0.029} & \textbf{0.026} & \textbf{0.086} & \textbf{0.024} & 0.006 & 0.015 & 0.015 & 0.003 & 0.005 & 0.011\\
1 & 0.015 & 0.011 & 0.014 & 0.010 & 0.003 & 0.004 & 0.012 & 0.015 & 0.006 & \textbf{0.014} & 0.009\\
2 & 0.018 & 0.007 & 0.011 & 0.013 & 0.005 & 0.010 & \textbf{0.026} & 0.018 & 0.007 & 0.005 & 0.013\\
3 & 0.013 & 0.007 & 0.003 & 0.007 & 0.007 & \textbf{0.013} & 0.016 & 0.017 & 0.005 & 0.009 & 0.011\\
4 & NA & NA & NA & NA & NA & NA & 0.017 & \textbf{0.026} & 0.005 & 0.008 & 0.005\\
5 & NA & NA & NA & NA & NA & NA & 0.016 & 0.020 & \textbf{0.008} & 0.008 & \textbf{0.040}\\
6 & NA & NA & NA & NA & NA & NA & 0.017 & 0.012 & 0.005 & 0.006 & 0.004\\
\addlinespace
12 & NA & NA & NA & NA & NA & NA & 0.010 & 0.013 & NA & NA & NA\\
\bottomrule
\end{tabular}}
\end{table}

% HERE ADD FIGURE OF GF RUN VARS FROM FINAL RUN NOT SHOWN IN FIGURE 6

>>>>

\journalname{JGR: Biogeosciences}

@@

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
\index Instead of a unidirectional trend, we see a shift around 2004 to low-upwelling conditions, and a return to the previous community and physical environment in 2014. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, but community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is sea surface temperature (SST), with the MEI v2 index (lag of 4 months) and nitrate concentration following behind. 
\end{keypoints}

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
Variable Name & Units & Data source \\
\midrule
MEI v.2 & unitless & Climate Indices provided by NOAA \\
AMO & unitless & Climate Indices provided by NSF NCAR \\
Wind speed & m/s & ERA5 model reanalysis: 10 m u-component of wind speed \\
Precipitation & 0.013 & ERA5 model reanalysis: total precipitation\\
Evaporation & NA & ERA5 model reanalysis: evaporation \\
SST & °C & CARIACO time series: CTD, Average temperature per month measured in top 10 m\\
21° C Isotherm & m & CARIACO time series: CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, top 100m \\
NO3 & µM & Niskin, top 100m \\
PO4 & µM & Niskin, top 100m \\
SiO4 & µM & Niskin, top 100m \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & **\\
Wind speed & 7074.0 & 0.143 & n.s.\\
Precipitation & 6251.0 & 0.004 & **\\
Evaporation & 9979.0 & 3.68e-5 & ***\\
\addlinespace
21 °C Isotherm & 5293.0 & 0.034 & *\\
SST & 4712.0 & 0.001 & ***\\
Salinity & 7636.0 & 0.002 & **\\
$NO_3$ & 6378.0 & 0.018 & *\\
$PO_4$ & 7359.0 & 0.012 & *\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
\addlinespace
Chlorophyll $a$ & 8229.0 & 9.63e-6 & ***\\
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & *\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
\caption{Gradient Forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
\label{fig:GF}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
\caption{Correlation matrix with clustering of environmental variables, diversity metrics and cell counts. Correlation was calcualted using a Spearman's rank correlation coefficient, the magnitude of the coefficient is indicated by the color of the circle. The size of the circle and significance level marking represents the calculated significance of the correlation deviating from 0 (p-value: ***\textless0.001, **\textless0.01, *\textless0.05). The time series were aditionally ordered by hierarchical clustering via the complete linkage method into 5 groups indicated by the black square border.}
\label{fig:sup:correlation}
\end{figure}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA2_GF_output_Supplement_v1.pdf}
\caption{Cumulative importance curves for all predictor variables not shown in the main figure. Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups. Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
\label{fig:sup:GFoutput_lags_extra}
\end{figure}

@@

\caption{Resulting $R^2$ weighted importance for running predictor variables together with time-lags to check if they provide a better fit to the data. Each column represents a single run, with resulting importance values for time series and monthly lags. The results were stable, so that repeated runs confirmed the chosen time lag. For in-situ variables,  we tested all measurements up to a lag of 3 months. For climate variables we tested up to a lag of 6 months and added a lag of 12 months for the climate indices. Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
lag & NO3 & PO4 & SiO4 & SST & \qty{21}{\celsius} Isotherm & Salinity & AMO & MEI v.2 & Wind speed & Precipitation & Evaporation\\
\midrule
0 & \textbf{0.027} & \textbf{0.029} & \textbf{0.026} & \textbf{0.086} & \textbf{0.024} & 0.006 & 0.015 & 0.015 & 0.003 & 0.005 & 0.011\\
1 & 0.015 & 0.011 & 0.014 & 0.010 & 0.003 & 0.004 & 0.012 & 0.015 & 0.006 & \textbf{0.014} & 0.009\\
2 & 0.018 & 0.007 & 0.011 & 0.013 & 0.005 & 0.010 & \textbf{0.026} & 0.018 & 0.007 & 0.005 & 0.013\\
3 & 0.013 & 0.007 & 0.003 & 0.007 & 0.007 & \textbf{0.013} & 0.016 & 0.017 & 0.005 & 0.009 & 0.011\\
4 & NA & NA & NA & NA & NA & NA & 0.017 & \textbf{0.026} & 0.005 & 0.008 & 0.005\\
5 & NA & NA & NA & NA & NA & NA & 0.016 & 0.020 & \textbf{0.008} & 0.008 & \textbf{0.040}\\
6 & NA & NA & NA & NA & NA & NA & 0.017 & 0.012 & 0.005 & 0.006 & 0.004\\
\addlinespace
12 & NA & NA & NA & NA & NA & NA & 0.010 & 0.013 & NA & NA & NA\\
\bottomrule
\end{tabular}}
\label{tab:sup:GFlagtests}
\end{table}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA3_GF_output_NOLAG_v1.pdf}
\caption{Gradient Forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
\label{fig:sup:GFoutput_nolags}
\end{figure}

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/FigureA4_GF_output_Supplement_NOLAG_v1.pdf}
\caption{Cumulative importance curves for all variables not shown in the previous figure. Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups. Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
\label{fig:sup:GFoutput_nolags_extra}
\end{figure}
